---
title: "PSTAT 131 Final Project"
author: "Kenneth Villatoro"
date: "2023-12-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE}
#libraries that will be used 
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(corrplot)
library(corrr)
library(MASS)
library(ISLR)
library(ISLR2)
library(ggthemes)
library(parsnip)
library(readxl)
library(ranger)
library(glmnet)
library(modeldata)
library(janitor)
library(naniar)
library(xgboost)
library(ranger)
library(vip)
library(corrplot)
tidymodels_prefer()
```

## Table of Contents ##

- Introduction
- Exploratory Data Analysis 
- Model Predictions 
- Conclusion



## Introduction ##

What I will be predicting in this final Project are the outcome of species in terms of the numeric predictors that the data set provides. The data set I will be using in this final project is the Palmer Penguins data set from Kaggle. The website of this data set is located below along with the code book in order to understand the values used during the prediction. :

Website: https://www.kaggle.com/datasets/parulpandey/palmer-archipelago-antarctica-penguin-data

Codebook: *sent separately as well*

- species: penguin species (Chinstrap, Adélie, or Gentoo)
- culmen_length_mm: culmen length (mm)
- culmen_depth_mm: culmen depth (mm)
- flipper_length_mm: flipper length (mm)
- body_mass_g: body mass (g)
- island: island name (Dream, Torgersen, or Biscoe) in the Palmer Archipelago (Antarctica)
- sex: penguin sex


## Explatory Data Analysis ##

First off, we read in the data set while making the variables species, island, and sex all factors. The reason for this is because these three variables are nominal variables, or in other words, categorical variables. These variables must be converted in order to use these variables for the prediction. We also took off any data with any missing values as it can alter the prediction by giving null values. 

```{r}

penguins<-read_excel("penguins_size.xlsx",
                       na = "NA")

penguins$species <- as.factor(penguins$species)
penguins$island <- as.factor(penguins$island)
penguins$sex <- as.factor(penguins$sex)

penguins <- na.omit(penguins)

head(penguins)

```

We then split the data into training and testing set. The proportion of data to be split will be 75 percent for the training set and 25 percent for the testing set. Stratified sampling based on the species variable in order to help predict based on that variable. The cross validation that will be used will be 5 folds that are stratified on the species variable as well. 

```{r}
#splitting into testing and training set using strata

set.seed(123)

penguins_split <- initial_split(penguins, prop = 0.75, strata = "species")

penguins_train <- training(penguins_split)
penguins_test <- testing(penguins_split)

penguins_folds <- vfold_cv(penguins_train, v = 5, strata = "species")

```



Below is a correlation matrix that will help determine any postive or negative correlations among the numeric variables. We see that there are postive correlations between flipper length and culmen length, body mass and culmen length, and flipper length and body mass. On the other hand, we see negative correlations between culmen length and culmen depth, flipper length and culmen depth, and body mass and culmen depth. 

```{r}
#correlation of penguins to determine relationships 

penguins_train %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot(type = "lower", diag = FALSE)

```

Going more in depth with the positive correlation, below is a plot analyzing the interaction between and flipper length and body mass based on the type of species. The observations on this plot is that the Chinstrap species contain more flipper length than the Gentoo species while having more body mass than the Adelie species. The body mass between the Gentoo relatively equal with the Chinstrap species while the flipper length between the Adelie and Chinstrap species are also relatively equal. 

```{r}
#plot to correlate differences with culmen depth and length based on species 


palette <- c("red", "blue", "green")

penguins_train %>% 
  ggplot(aes(x = culmen_depth_mm, y = culmen_length_mm, 
             color = species)) +
  geom_point() +
  scale_color_manual(values = palette) +
  theme_minimal() +
  xlab("Flipper Length (mm)") +
  ylab("Body Mass (g)") + 
  labs(color = "Species")


```


In addition, sex must also be accounted for in order to predict the type of species the penguins are. Below is a plot that helps determine the relationship between Sex and a numeric variable, in this case will be body mass. We analyze that the Gentoo species between Male and Female contain higher body mass than either Adelie and Chinstrap species. Adelie and Chinstrap species are more equal to each other, with the Chinstrap female penguins having more penguins with higher body mass than Adelie female penguins. It is more even in the male plot with observations spread out. 

```{r}
#plot to correlate differences between sex and body mass based on species 


penguins_train %>% 
  ggplot(aes(x = sex, y = body_mass_g, 
             color = species)) +
  geom_point() +
  scale_color_manual(values = palette) +
  theme_minimal() +
  xlab("Flipper Length (mm)") +
  ylab("Body Mass (g)") + 
  labs(color = "Species")

```


## Model Predictions ##

Having all of these positive and negative correlations taken into account, the recipe that will be used will be predicting species using all numeric variables. Categorical variables will be dummy variables and the recipe will have interactions between culmen length (mm) and culmen depth (mm), and flipper length (mm) and body mass (grams). All predictors will also be centered and scaled. 

```{r}
#recipe for prediction 

penguins_recipe <- recipe(species ~ island + culmen_length_mm + culmen_depth_mm
                          + flipper_length_mm + body_mass_g + sex,
                          data = penguins_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_interact(terms = ~starts_with("culmen_length_mm"):culmen_depth_mm) %>%
  step_interact(terms = ~flipper_length_mm:body_mass_g) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

```


The four models that will be used to conduct the predictions are random forest, elastic net **logistic** regression, knn, and gradient boost trees. For the random forest, mtry, trees, and min_n will be tuned. For Elastic net logistic regression, mixture and penalty will be tuned. For knn, only neighbors will be tuned. Finally, for Gradient Boost Trees, mtry, trees, and the learn rate will be tuned. Below is the set up for the engines, mods, and workflows for all four models. 

```{r}
#models to be used to predict species: random forest, 
#elastic net **logistic** regression, knn, and gradient boost trees 

#random forest mod and workflow 
rf_class <- rand_forest(mtry = tune(), 
                           trees = tune(), 
                           min_n = tune()) %>%
  set_engine("ranger") %>% 
  set_mode("classification")

rf_class_wf <- workflow() %>% 
  add_model(rf_class) %>% 
  add_recipe(penguins_recipe)

#elastic mod and workflow
elastic_mod_class <- multinom_reg(mixture = tune(),penalty = tune()) %>%
  set_mode("classification") %>%
  set_engine("glmnet")

elastic_wkflow_class <- workflow() %>%
  add_model(elastic_mod_class) %>%
  add_recipe(penguins_recipe)


#knn mod and workflow 
knn_mod_class <- nearest_neighbor(neighbors = tune()) %>%
  set_mode("classification") %>%
  set_engine("kknn")

knn_wkflow_class <- workflow() %>% 
  add_model(knn_mod_class) %>% 
  add_recipe(penguins_recipe)


#gradient boost trees 
bt_class_spec <- boost_tree(mtry = tune(), 
                           trees = tune(), 
                           learn_rate = tune()) %>%
  set_engine("xgboost") %>% 
  set_mode("classification")

bt_class_wf <- workflow() %>% 
  add_model(bt_class_spec) %>% 
  add_recipe(penguins_recipe)


```


Below are the grid setups. For knn, the neighbors will be from a range of 1 to 10, conducting 10 levels of these grid. For Elastic Net, the penalty will be set to default, mixture will be from range 0 to 1, while conducting 10 levels in the grid. For random forest, mtry will be from 1 to 6 as that is the maximum range of the variables being analyzed, a range of 100 to 400 trees, min_n from a range of 10 to 20, while conducting 5 levels of these predictions. Finally, for Gradient Boost Trees, mtry will also be from 1 to 6 and 50 to 300 trees. The learn rate will be from a range of -10 to -1, with 5 levels of these predictions in the grid. 

```{r}
#grid and tuning 

#grids
knn_grid <- grid_regular(neighbors(range= c(1,10)), levels = 10)

elastic_grid <- grid_regular(penalty(), mixture(range = c(0,1)), levels = 10)

rf_grid <- grid_regular(mtry(range = c(1, 6)), 
                        trees(range = c(100, 400)),
                        min_n(range = c(10, 20)),
                        levels = 5)

bt_grid <- grid_regular(mtry(range = c(1, 6)), 
                        trees(range = c(50, 300)),
                        learn_rate(range = c(-10, -1)),
                        levels = 5)

```


After creating the grids for each model, each model must be tuned using the workflows created prior, while adding the grids to each workflow and resamples from the cross validation in the beginning of the prediction process. 

```{r, eval =FALSE}
#tuning and collecting the metrics for each model and fitting to final workflow

#tuning

tune_knn_class <- tune_grid(object = knn_wkflow_class,
                      grid = knn_grid,
                      resamples = penguins_folds,
                      )

tune_elastic_class <- tune_grid(object = elastic_wkflow_class,
                          grid = elastic_grid,
                          resamples = penguins_folds,
                          )

tune_rf_class <- tune_grid(rf_class_wf,
                           resamples = penguins_folds,
                           grid = rf_grid)

tune_bt_class <- tune_grid(bt_class_wf,
                           resamples = penguins_folds,
                           grid = bt_grid)
save(tune_knn_class, file = "knn_class.rda")
save(tune_elastic_class, file = "elastic_class.rda")
save(tune_rf_class, file = "rf_class.rda")
save(tune_bt_class, file ="bt_class.rda")
```


Below are the roc-auc and accuracy metrics that are collected from each model after tuning the workflows:

```{r}
load("knn_class.rda")
load("elastic_class.rda")
load("rf_class.rda")
load("bt_class.rda")

#metrics for each model
collect_metrics(tune_knn_class)
collect_metrics(tune_elastic_class)
collect_metrics(tune_rf_class)
collect_metrics(tune_bt_class)
```


Below are the best models that were conducting after tuning the workflows using the grids and resamplings from the cross validation. These best values will be used in order to finalize the workflow for each model. 

```{r, warning=FALSE}
show_best(tune_knn_class)
show_best(tune_elastic_class)
show_best(tune_rf_class)
show_best(tune_bt_class)
```

After storing each best models in its each indivudal assignment, it is imperative we make sure that the metric being used is only roc-auc. Otherwise, R will be confused in trying to figure out with metric to use in order to finalize the workflow. We then finalize out workflows for each model shown below:

```{r}
#best model of each tuning 

best_knn <- select_best(tune_knn_class, metric = "roc_auc")
best_elastic<- select_best(tune_elastic_class, metric = "roc_auc")
best_rf <- select_best(tune_rf_class, metric = "roc_auc")
best_bt <- select_best(tune_bt_class, metric = "roc_auc")

#finalizing workflows 
final_wkflow_knn <- finalize_workflow(knn_wkflow_class, best_knn)
final_wkflow_elastic <- finalize_workflow(elastic_wkflow_class, best_elastic)
final_wkflow_rf <- finalize_workflow(rf_class_wf, best_rf)
final_wkflow_bt <- finalize_workflow(bt_class_wf, best_bt)

```

After finalizing the workflows, the workflows are then fitted into the training dataset: 

```{r}
#fitting workflows

fitted_final_wkflow_knn <- fit(final_wkflow_knn, data = penguins_train)
fitted_final_wkflow_elastic <- fit(final_wkflow_elastic, data = penguins_train)

fitted_final_wkflow_rf <- fit(final_wkflow_rf, data = penguins_train)
fitted_final_wkflow_bt <- fit(final_wkflow_bt, data = penguins_train)

```


After fitting the finalized workflows, they will finally be fitted now using the testing dataset in order to determine which model or two models predicting the outcome of species the best. 

```{r}
#fitting all models into testing set and choosing best one or two models 

testing_bt <- augment(fitted_final_wkflow_bt, new_data = penguins_test) %>%
    roc_auc(truth = species, .pred_Adelie:.pred_Gentoo)

testing_elastic<-augment(fitted_final_wkflow_elastic, new_data = penguins_test) %>%
  roc_auc(truth = species, .pred_Adelie:.pred_Gentoo)

testing_knn <- augment(fitted_final_wkflow_knn, new_data = penguins_test) %>%
  roc_auc(truth = species, .pred_Adelie:.pred_Gentoo)

testing_rf <- augment(fitted_final_wkflow_rf, new_data = penguins_test) %>%
  roc_auc(truth = species, .pred_Adelie:.pred_Gentoo)

testing_bt
testing_elastic
testing_knn
testing_rf
```


Conclusion

- We see that the the ROC-AUC for Random Forest, Elastic Net Logistic Regression, and Boost Trees had no problem in predicting the species outcome of the penguins as it was 1. KNN also had a great score but was the least with a score of 0.993. I am honestly not surpised that the models did really well as the postive interactions played a significant role in determining which observations were part of an specific species. I would suggest doing Neural Networks and SVMs in order to further evaluate the data set and whether or not having different interactions of the data set would effect the outcome of the machine learning algorithms. In addition, this data set can also be analyzed through other aspects such as body mass, type_1, and other variables that the data set can offered, which is also another alternative in helping further evaluate machine learning algorithms.


